\textbf{Appendix}\\ Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-08-21 Chapter_0/04_Manuscript/Manuscript_Backup_2020-01-12.tex
first commit

% Title of the document
{\textsc{ \LARGE Running Wild: Movement Corridors of Dispersing African Wild
Dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation
Area}}\\[0.4cm]

@@

\begin{abstract}
The African wild dog \textit{(Lycaon pictus)} is an endangered species that
currently only survives in a few scattered subpopulations in Sub-Saharan Africa.
Long-term viability of the species thus requires the identification and
preservation of key dispersal corridors that connect such subpopulations. We
collected GPS data of 16 dispersing coalitions originating from a free-ranging
population in northern Botswana and used integrated step selection analysis to
create a permeability surface spanning over the entire Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA), the largest transboundary
conservation area on earth. Based on our permeability surface, we calculated
least-cost paths and least-cost corridors between selected locations within the
KAZA-TFCA. Our results show that dispersal corridors typically run through open
shrubs/grassland, and in the vicinity of water. Corridors rarely cross
waterbodies, areas with high human presence, or areas densely vegetated by
trees. Several dispersal corridors meet in the Linyanti-Chobe ecosystem, which
highlights the importance of northern Botswana for the conservation of this
charismatic species.
\end{abstract}
\newpage

@@

\section{Introduction}
% Importance of Connectivity
Successful conservation of spatially fragmented subpopulations requires to
assess connectivity of habitats occupied by these subpopulations. In this
context, connectivity describes the degree to which the landscape facilitates or
impedes movement of the focal species between habitat patches
\citep{Taylor.1993, Clobert.2012}. Thus, connectivity is a prerequisite for
dispersal, which in turn promotes genetic exchange, reinforcement of small
populations, and colonization or recolonization of unoccupied habitats
\citep{Brown.1977, Hanski.1998, MacArthur.2001, Frankham.2002, Leigh.2012}. All
of these processes crucially contribute to the long-term viability of numerous
species, therefore emphasizing the importance of connectivity. However,
increasing anthropogenic pressure and ongoing loss and fragmentation of valuable
habitats erode connectivity on a worldwide scale \citep{Fahrig.2003,
Barnosky.2012}. As a result, management actions for the conservation of
endangered species increasingly aim to restore connectivity through the
identification and preservation of key movement corridors that link isolated
habitats \citep{Fahrig.2003, Heller.2009}. An unprecedented instance of an
initiative to restore connectivity on a large scale is the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA), which traverses five African
countries and spans over 520,000km\textsuperscript{2}. While the primary purpose
of the KAZA-TFCA is to facilitate the migration of elephants across large
landscapes, numerous other wide-ranging species will likely benefit from its
planned protectorates too. Nevertheless, the extent to which such initiatives
are beneficial to certain species is rarely assessed, leaving open their true
impact on connectivity.

% African wild dogs
One of the species for which the degree to which it will benefit from the
KAZA-TFCA remains unknown is the African wild dog (\textit{Lycaon pictus}), a
keystone predator and flagship species for conservation efforts in the KAZA-TFCA
\citep{VanDerMeer.2016}. While once present across the entire Sub-Saharan
continent, the species has been widely extirpated through human persecution,
habitat destruction and disease outbreaks. As a result, the African wild dog has
become Africa's most endangered large carnivore that currently only survives in
small, spatially scattered subpopulations (see \Cref{StudyArea}a,
\cite{Woodroffe.2012}). Within these subpopulations, wild dogs form cooperative
breeding groups of up to thirty individuals in which a single dominant pair
monopolizes the majority of reproduction \citep{Frame.1979, Fuller.1992,
Creel.2002}, while subordinates of both sexes help to feed newborn pups
\citep{Malcolm.1982}. After reaching 15 months of age, male and female offspring
disperse from their natal pack in same-sex coalitions and search for unrelated
mates and a suitable territory to settle (\citeauthor{McNutt.1996}, 1996;
\citeauthor{Behr.2019}, unpublished). During such dispersal events, wild dogs
cover vast distances, sometimes crossing perilous areas that are strongly
influenced by humans (\citeauthor{DaviesMostert.2012}, 2012;
\citeauthor{Masenga.2016}, 2016; \citeauthor{Cozzi.2019}, in review). Despite
the importance of dispersal for the remaining wild dog subpopulations, little is
known about the movement corridors of dispersing individuals and connectivity
between subpopulations in the KAZA-TFCA in general.

% Connectivity Assessment
In recent years, there has been a growing body of literature that uses animal
movement data to identify movement corridors and thereby assess connectivity on
large scales \citep{Chetkiewicz.2006}. Revelation of potential movement
corridors between selected locations traditionally relies on the estimation of a
permeability surface (elsewhere referred to as resistance surface when
inverted), which depicts the ease or willingness at which the focal species
traverses a specific landscape. Latest and very powerful empirical methods to
predict permeability based on observed movements include \textit{resource
selection functions} (RSF, \cite{Boyce.2002}), \textit{step selection functions}
(SSF, \cite{Fortin.2005}) and \textit{path selection functions} (PSF,
\cite{Cushman.2010}). All methods have in common that they infer habitat
preferences of the focal species by comparing environmental covariates at
visited locations to environmental covariates at randomly selected locations
that are deemed available but presumably unused by the animal. Based on inferred
habitat preferences, the methods further assume that habitats preferred by the
focal species provide high permeability, whereas avoided habitats provide low
permeability. Regardless of the method chosen (i.e. RSF, SSF, or PSF), all
approaches require adequate relocation data that is representative for the
process being studied.

% Behavioral Modes
For connectivity studies in particular, the behavioral mode of the animal (e.g.
resident, dispersing) during data collection should be considered when
estimating landscape permeability and assessing movement corridors. However, few
studies have partitioned their movement data according to behavioral mode
\citep{Wilson.2012, Vasudev.2015}. As a consequence, most researchers estimated
permeability for dispersing individuals as if they had the same preferences as
resident individuals. This is a rather disputed assumption and recent research
suggests that such an approach seriously reduces the power to reveal meaningful
movement corridors \citep{Killeen.2014}. For instance, \cite{Elliot.2014}
demonstrated marked differences in predicted permeability surfaces for
dispersing lions in the KAZA-TFCA depending on whether permeability was assessed
using data collected during residence or dispersal. \cite{Jackson.2016} made
similar observations for African wild dogs and revealed higher permeability for
dispersers than for residents. Moreover, \cite{Abrahms.2017} found a
significantly greater overlap between predicted and observed corridors for three
dispersing wild dogs when predictions were based on information collected during
wild dog activity, rather then when ignoring the behavioral mode. Overall,
predicted movement corridors vastly differed depending on the type of data that
was being used \citep{Elliot.2014, Jackson.2016, Abrahms.2017}. Unfortunately,
data of dispersing wild dogs remains scarce and is collected opportunistically
at best \citep{DaviesMostert.2012, Masenga.2016, Jackson.2016, Abrahms.2017}.
For instance, the dispersers reported in \cite{DaviesMostert.2012} were not
fitted with global positioning system (GPS) collars, which is why the
researchers only recorded straight line dispersal-distances.
\cite{Masenga.2016}, on the other hand, were able to collect proper GPS data,
yet only on two dispersers. Furthermore, \cite{Jackson.2016} based their work on
occasional wild dog sightings and their distinction between residents and
dispersers was based purely on group size, assuming that any group with fewer
than nine members was dispersing. Finally, \cite{Abrahms.2017} used GPS
relocation data of resident individuals during phases of activity rather than
during true dispersal. In summary, none of these studies systematically
monitored dispersing individuals, which is likely owed to technological and
logistic limitations when monitoring such an elusive and wide-ranging species
\citep{Nathan.2001, Tesson.2013}. Yet, a better understanding of how the
landscape impedes or facilitates movement of dispersers is vital for improving
our understanding of connectivity and for conservation of the species in
general.

% Research Goal
Our study therefore had two aims: (1) to estimate landscape permeability for
dispersing wild dogs, and subsequently use this information to (2) assess
connectivity by revealing key dispersal corridors within the entire extent of
the KAZA-TFCA. For this purpose, we collected GPS data of 16 dispersing
coalitions originating from a free-ranging population of wild dogs in the
eastern section of the Okavango Delta in Botswana. We highlight key dispersal
corridors that provide connectivity between already established protected areas
within the KAZA-TFCA and we reveal potentially important dispersal corridors
outside the planned protectorates of the KAZA-TFCA. Our results have important
implications for the conservation and management of the endangered African wild
dog.

\newpage
\section{Methods}
\subsection{Study Area}
% Core Study Area
For the purpose of this study we differentiated between a \textit{core study
area}, where all dispersers originated from, and an \textit{extended study
area}, for which we conducted the connectivity analysis. The \textit{core study
area} was located in the eastern section of the Okavango Delta in northern
Botswana (\Cref{StudyArea}b, centered at -19\degree 25'30''S, 23\degree
37'30''E; elevation ca. 950m) and comprised parts of the Moremi Game Reserve and
adjacent wildlife management areas. The Okavango Delta is considered to be one
of the main strongholds of African wild dogs in southern Africa and likely acts
as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012, Cozzi.2013}. The Okavango Delta represents the earth's
largest inland delta and is characterized by diverse habitats, ranging from
permanent river systems, mopane woodland and acacia forests, to open grasslands
and seasonal swamps \citep{Broekhuis.2013}. Rainfall is periodic and lasts from
November to March, whereas the rest of the year remains dry \citep{McNutt.1996}.
Flooding of the delta greatly changes depending on the season and drastically
affects surrounding landscapes (see \Cref{FloodPulse} in
\Cref{Appendix:FloodPulse} for an illustration of a typical flood pulse). During
maximum extent, the delta becomes a patchy conglomerate of swamps, open water,
and islands. Because rainfalls only slowly descent from their catchment areas in
southern Angola into the Okavango Delta, the flood usually peaks at the height
of the dry season between August and September \citep{Wolski.2017}. Most of the
Okavango Delta's unique basin is under protection, such that human influence in
our core study area was mainly restricted to the surroundings of Maun and the
smaller villages of Khwai, Mababe, and Sankuyo.

% Large Study Area
Our \textit{extended study area} was confined by the extent of the KAZA-TFCA
(\Cref{StudyArea}a, centered at -17\degree 13'9''S, 23\degree 56'4''E; elevation
ca. 1,030m), which lies in the basins of the Okavango and Zambezi rivers and
embraces parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. With a total
area of over 520,000km\textsuperscript{2} the KAZA-TFCA contains several already
existing national parks (NPs) and other protected areas, and will eventually
become earth's largest transboundary conservation area. Since we considered the
entire rectangular extent of the KAZA-TFCA, our extended study area encompassed
a total of over 1.3 Mio km\textsuperscript{2}. Similar to our core study area,
the extended study area hosts a conglomerate of various landscapes, including
savanna, grassland, and dry or moist woodland patches. Furthermore, the
KAZA-TFCA encloses several wild dog subpopulations  that still reside in
Namibia, Botswana, and Zimbabwe (\Cref{StudyArea}a, \cite{Woodroffe.2012}).
Despite 36 NPs and numerous other protected areas, there is considerable human
influence in some regions of the KAZA-TFCA originating from farms, high human
density, and road traffic.

@@

\subsection{GPS Relocation Data}
% Collection
We collected GPS relocation data of 16 wild dog coalitions (i.e. 7 female and 9
male coalitions) that dispersed from the core study area between 2011 and 2019.
Potential dispersing individuals were immobilized according to protocols
described in \cite{Osofsky.1996} and fitted with GPS/Satellite radio collars
(\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while they were still
with their natal pack. All required procedures were undertaken and supervised by
a Botswana-registered wildlife veterinarian. Immobilized individuals were
monitored and all were observed to successfully re-join their packs within one
hour after immobilization. GPS collars were programmed to record and transmit
GPS relocations via Iridium satellite system to a base station. The system thus
allowed remote-tracking of collared individuals, even where field conditions
were prohibitive or dispersal trajectories crossed international borders.
Because we were mainly interested in recording dispersal trajectories,
relocations were collected only every 24 hours during residence, but every 4
hours during dispersal. Dispersal and residence were strictly distinguished
using a combination of net squared displacement \citep{Borger.2012}, visual
inspection of trajectories, and observations of pack constellations in the field
(for details see \citeauthor{Cozzi.2019}, in review). Dispersal was deemed to
have started when an individual left its natal home range, and continued until
the individual formed a new pack or exhibited signs of resident behavior. Some
dispersers undertook exploratory movements prior to dispersal. Such movements
have been observed in other species as well and are reported to be very similar
to true dispersal \citep{Killeen.2014}. We therefore classified exploratory
movements as dispersal too. We did not attempt to distinguish sex in our
analysis, as previous research found little differences between males and
females during dispersal (\citeauthor{Woodroffe.2019}, 2019;
\citeauthor{Cozzi.2019}, in review). In total, we collected 4,691 GPS
relocations during dispersal (see \Cref{Trajectories}). On average, dispersal
events included three individuals (min = 1, max = 7), lasted for 41 days (min =
7, max = 133), and covered a mean cumulative distance of 370km (min = 105, max =
610). Detailed descriptions of the dispersal events can be found in
\citeauthor{Cozzi.2019} (in review). Although our final sample size of 16
dispersers was relatively small, we consider this an exceptional number for such
an endangered and elusive species.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Trajectories.pdf}
    \caption{Illustration of the dispersal trajectories that we recorded. Each
    color represents a different dispersal coalition. All coalitions departed
    from our core study area, which is indicated by the white square. The
    coalition dispersing towards the far east of the map covered over 360km in
    under 10 days.}
    \label{Trajectories}
  \end{center}
\end{figure}

\subsection{Covariates}
% General stuff
We used a set of geo-referenced covariates to investigate movement and habitat
preferences of dispersing wild dogs (see \Cref{Covariates}). Covariates fell
into one of the three categories \textit{land cover}, \textit{protection
status}, and \textit{human influence}. For each category we obtained spatial
raster layers from freely available online services or from remote sensed
satellite imagery. To ensure a consistent resolution (i.e. cell-size or grain)
across covariates, all layers were coarsened or interpolated to match a
resolution of 250m x 250m. Processing and manipulation of data, as well as all
spatial and statistical analyses were performed in R, version 3.6.1
\citep{R.2019}.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Covariates.pdf}
    \caption{Overview of covariates that we used in our models. Although all
    covariates were prepared for the extended study area, we present only an
    extract to maximize visibility. The white square in each plot depicts our
    core study area. (a) Averaged layer of all binary maps of water and dryland
    that we acquired using our remote sensing algorithm. (b) Percentage cover of
    trees. (c) Percentage cover of shrubs/grassland. Note that anything not
    covered by trees or shrubs/grassland was deemed to be bareland. (d)
    Protection status according to the Peace Parks Foundation. (e) Human
    influence proxy, including human density, farms, and roads. (f) Distance to
    nearest road (white lines depict actual roads).}
    \label{Covariates}
  \end{center}
\end{figure}

\subsubsection{Land Cover}
% Overview
Land cover was represented by four layers: (1) a binary layer distinguishing
water and dryland, and (2) three continuous layers showing the percentage cover
by different vegetation types, together adding up to 100\%.

% Dynamic Watermapping
(1) For the purpose of this study, water included open waters, rivers, wetlands,
and swamps, whereas anything else was presumed to be dryland. Because flooding
of the Okavango Delta is highly variable, we decided to map the flood
dynamically for the extent shown in \Cref{Floodmapping}. To create dynamic
``floodmaps'', we applied an algorithm that was originally designed by the
\textit{Okavango Research Institute} (ORI) and is explained in detail by
\cite{Wolski.2017}. However, we rewrote and slightly amended the original Python
code in R and thus describe the underlying principles again and especially
indicate the discrepancies to the original algorithm.

% Algorithm
To implement the algorithm, we defined two sets of polygons located in the
region of the Okavango Delta (see \Cref{Floodmapping}). The first set consisted
of areas known to be permanent dryland, whereas the second set consisted of
permanent waters. Since we were unable to retrieve the original polygons used in
\cite{Wolski.2017}, we geo-referenced one of their illustrations and traced the
polygons there. After recreation of the polygons, we used the R-package
\textit{getSpatialData} \citep{Schwalb.2018} to download and pre-process all
relatively cloud-free MODIS Terra images (MCD43A4, \cite{Schaaf.2015}) available
for the period of our dispersal events. Assessment of cloud cover was based on
visual inspection of MODIS images on ORI's website \citep{ORI.2019}. The MCD43A4
MODIS product is particularly useful for dynamic floodmapping because it is
updated with new composite images every 8th day \citep{Wolski.2017}. After
download, we classified each MODIS image into a binary map of water (flood) and
dryland using a threshold that was identified as follows. First, we extracted
all reflectance values of MODIS Terra Band 7 within the water- and
dryland-polygons. Second, we computed histograms of water-reflectances and
dryland-reflectances and empirically verified that reflectances of the two
groups were sufficiently distinct. More specifically, we checked if
superimposing the two histograms resulted in a bimodal plot. This was said to be
achieved if the 99\textsuperscript{th} percentile of water-reflectances did not
severely exceed the 1\textsuperscript{st} percentile of dryland-reflectances
(\(p_{0.99, water} - \frac{10}{255} < p_{0.01, dryland}\)). Third, if peaks were
sufficiently different, we calculated a threshold (\(t\)) using \Cref{EQ1}:

>>>>

% Title of the document
{\textsc{ \LARGE Ranging Wide: Movement Corridors of Dispersing African Wild
Dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation
Area}}\\[0.4cm]

@@

\begin{abstract}
The African wild dog \textit{(Lycaon pictus)} is an endangered species that
currently only survives in a few scattered subpopulations in Sub-Saharan Africa.
Long-term viability of the species thus requires the identification and
preservation of key dispersal corridors that connect such subpopulations. We
collected GPS data of 16 dispersing coalitions originating from a free-ranging
population in northern Botswana and used integrated step selection analysis to
create a permeability surface spanning over the entire Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA), the largest transboundary
conservation area on earth. Based on our permeability surface, we calculated
least-cost paths and least-cost corridors between selected locations within the
KAZA-TFCA. Our results show that dispersal corridors typically run through open
shrubs/grassland and in the vicinity of water. Corridors rarely cross
waterbodies, areas with high human presence, or areas densely vegetated by
trees. Several dispersal corridors meet in the Okavango-Linyanty ecosystem,
which highlights the importance of northern Botswana for the conservation of
this charismatic species.
\end{abstract}
\newpage

@@

\section{Introduction}
% Importance of Connectivity
Successful conservation of fragmented and isolated subpopulations requires
maintenance and enhancement of connectivity between habitats occupied by these
subpopulations \citep{Hanski.1998}. In this context, connectivity describes the
degree to which the landscape matrix facilitates or impedes movement of
individuals and thereby genes of the focal species between habitat patches
\citep{Taylor.1993, Clobert.2012}. Thus, sufficient connectivity facilitates
dispersal, which in turn promotes genetic exchange, reinforcement and rescuing
of small populations, and colonization or recolonization of unoccupied habitats
\citep{Brown.1977, Hanski.1998, MacArthur.2001, Frankham.2002, Leigh.2012}. All
these processes crucially contribute to long-term population viability, which
emphasizes the importance of dispersal and connectivity. However, increasing
anthropogenic pressure and ongoing habitat loss and fragmentation reduce
connectivity on a worldwide scale \citep{Fahrig.2003, Barnosky.2012}. As a
result, management actions for the conservation of endangered species are
increasingly aimed towards preserving and restoring connectivity through the
identification and preservation of key dispersal corridors that link isolated
habitats \citep{Fahrig.2003, Heller.2009}. An unprecedented initiative to
restore connectivity on a large scale is represented by the creation of the
Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), a protected area
which traverses five African countries and spans over 520,000
km\textsuperscript{2}. While such initiatives are expected to bring benefits to
wide-ranging species such as elephants (\textit{Loxodonta africana}), African
wild dogs (\textit{Lycaon pictus}), and lions (\textit{Panthera leo}), the
extent to which they are effective for targeted species and whether boundaries
reflect natural dispersal corridors is rarely assessed.

% Connectivity Assessment
In recent years, there has been a growing body of literature that uses animal
movement data to identify dispersal corridors and to assess connectivity on
large scales \citep{Chetkiewicz.2006}. Identification of potential dispersal
corridors between suitable habitat patches traditionally relies on the
estimation of a permeability surface (elsewhere referred to as resistance
surface, when inverted), which renders the ease or willingness at which the
focal species traverses a specific landscape. Latest empirical methods to
predict landscape permeability based on recorded GPS relocation data include
\textit{resource selection functions} (RSF, \cite{Boyce.2002}), \textit{step
selection functions} (SSF, \cite{Fortin.2005}), and \textit{path selection
functions} (PSF, \cite{Cushman.2010}). All methods have in common that they
infer habitat preferences of the focal species by comparing covariates (e.g.
environmental and anthropogenic) at visited locations to the same covariates at
randomly selected locations that are deemed available but presumably unused by
the animal. Based on inferred habitat preferences the methods further assume
that habitats preferred by the focal species provide high permeability, whereas
avoided habitats provide low permeability. Regardless of the method chosen (i.e.
RSF, SSF, or PSF), all approaches require adequate GPS relocation data that is
representative for the process being studied. More specifically, it has been
shown that GPS relocation data collected on dispersing individuals is better
suited for detecting dispersal corridors than GPS relocation data collected on
resident individuals \citep{Elliot.2014}. Few studies have, however, partitioned
their GPS relocation data according to behavioral mode such as \textit{resident}
vs. \textit{dispersing} or \textit{active} vs. \textit{inactive} periods of
movement \citep{Wilson.2012, Vasudev.2015}. As a consequence, most permeability
surfaces upon which dispersal corridors are identified are created using GPS
relocation data that was collected on resident individuals. This introduces
severe biases and substantially reduces the power to reveal meaningful
corridors, for dispersing individuals have different needs and motivations
compared to resident individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}.
For instance, \cite{Elliot.2014} demonstrated marked differences in predicted
permeability surfaces for dispersing lions in the KAZA-TFCA depending on whether
permeability was assessed using data collected during residence or dispersal.
\cite{Jackson.2016} made similar observations for African wild dogs and revealed
higher permeability for dispersers than for residents. Moreover,
\cite{Abrahms.2017} found a significantly greater overlap between predicted and
observed corridors for three dispersing wild dogs when predictions were based on
information collected only during wild dog activity, rather than across the
entire day.

% African wild dogs
A highly endangered flagship species for conservation efforts across the
KAZA-TFCA, but for which systematic GPS relocation data on dispersing
individuals is largely unavailable, is the African wild dog
\citep{VanDerMeer.2016}. While once present across the entire Sub-Saharan
continent, the species has been widely extirpated through human persecution,
habitat destruction, and disease outbreaks \citep{Woodroffe.2012}. As a result,
the African wild dog has become one of Africa's most endangered large carnivores
and currently only survives in small, spatially scattered subpopulations
\citep{Woodroffe.2012}. Within these subpopulations, wild dogs form cooperative
breeding groups of up to thirty individuals in which a single dominant pair
monopolizes the majority of reproduction \citep{Frame.1979, Fuller.1992,
Creel.2002}, while subordinates of both sexes help to feed newborn pups
\citep{Malcolm.1982}. After reaching 15 months of age, male and female offspring
disperse from their natal pack in same-sex coalitions and search for unrelated
mates and a suitable territory to settle (\citeauthor{McNutt.1996}, 1996;
\citeauthor{Behr.2019}, unpublished). During such dispersal events, wild dogs
are able to cover several hundred kilometers, sometimes crossing risky areas
that are strongly influenced by humans \citep{DaviesMostert.2012, Masenga.2016,
Cozzi.2020}. Despite the importance of dispersal for the remaining wild dog
subpopulations, little is known about the corridors of dispersing individuals
and connectivity between subpopulations in the KAZA-TFCA in general.

% Covariate Motivation
Wild dog movements and spatial behavior are known to be influenced by a suite of
environmental and anthropogenic factors. For instance, \cite{Abrahms.2017} found
that wild dog movements during residence mainly depended on land cover,
protection status and anthropogenic features. Similar observations on resident
packs were made by \cite{Pretorius.2019}, who's results indicate that pack
presence was contingent on factors such as agricultural activity, road presence,
and protection status of the area. With respect to dispersing individuals
\cite{Whittington.2011} demonstrated that elevation, land cover, road density,
and human density best explained the distribution of recorded sightings in South
Africa. Also based on data from dispersing wild dogs, \cite{Jackson.2016} showed
that land cover, human density and road density significantly influenced the
probability of occurrence of dispersing wild dogs. Finally, \cite{Masenga.2016}
reported a low willingness of dispersers to cross human-dominated landscapes.
Large scale spatial information on the above mentioned factors can be obtained
either through freely available online services or through remote sensed
satellite imagery.

% Our Study
We used GPS relocation data collected on 16 wild dogs in as many dispersing
coalitions from a free ranging population in northern Botswana to estimate the
degree of selection or avoidance for environmental and anthropogenic landscape
features. We then used the obtained information in combination with relevant
spatial covariates to create a permeability surface for the entire KAZA-TFCA.
Lastly, we used the permeability surface as input for the computation of
least-cost paths and least-cost corridors between selected locations. This
allowed us to highlight key dispersal corridors within and outside the
boundaries of KAZA-TFCA and concluded with the identification of a ``dispersal
hub'' and of dispersal corridors that run outside formally protected areas.

\newpage
\section{Methods}
\subsection{Study Area}
% Both Study Areas
We differentiated between a \textit{core study area}, where all dispersers
originated from, and an \textit{extended study area}, which was represented by a
bounding box encompassing the KAZA-TFCA \citep{Cozzi.2020} and for which we
calculated connectivity.

% Core Study Area
The \textit{core study area} was located in the eastern section of the Okavango
Delta in northern Botswana (centered at -19\degree 25'30''S, 23\degree 37'30''E;
elevation ca. 950 m, orange rectangle in \Cref{StudyArea}b) and comprised parts
of the Moremi Game Reserve and adjacent wildlife management areas. The Okavango
Delta is one of the main strongholds of African wild dogs in southern Africa and
likely acts as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012, Cozzi.2013}. It is the earth's largest inland delta and
is characterized by diverse habitats, ranging from permanent river systems,
mopane woodland, and acacia forests, to open grasslands and seasonal swamps
\citep{Broekhuis.2013}. Rainfall is seasonal and lasts from November to March,
whereas the remainder of the year remains dry \citep{McNutt.1996}. The extent of
the flood greatly changes within and between years depending on the amount of
rain that descends from the catchment areas in Angola and reaches the distal
ends of the delta between July and August (see \Cref{FloodPulse} in
\Cref{Appendix:FloodmappingAlgorithm} for an illustration of a typical flood
pulse). The flood thereby drastically affects surrounding landscapes, so that
during maximum extent (ca. 12'000 km\textsuperscript{2}) the delta becomes a
patchy conglomerate of swamps, open water, and islands, whereas it shrinks
substantially (ca. 5'000 km\textsuperscript{2}) during dry periods
\citep{Gumbricht.2004, Wolski.2017}. Most of the Okavango Delta's unique basin
is under protection, such that human influence in our core study area was mainly
restricted to the surroundings of Maun and the smaller villages of Khwai,
Mababe, and Sankuyo.

% Large Study Area
The \textit{extended study area} (centered at -17\degree 13'9''S, 23\degree
56'4''E; elevation ca. 1,030 m, red rectangle in \Cref{StudyArea}a) was confined
by a bounding box encompassing the entire KAZA-TFCA. The KAZA-TFCA lies in the
basins of the Okavango and Zambezi rivers and embraces parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia. With a total area of over 520,000
km\textsuperscript{2} it contains several already-existing national parks (NPs)
and other protected areas. It constitutes the earth's largest transboundary
conservation area, which resulted in a rectangular study area of over 1.3 Mio
km\textsuperscript{2}. Similar to the core study area, the extended study area
hosts various landscapes, including savanna, grassland, and dry or moist
woodland habitats. Despite 36 NPs and other protected areas, there is
considerable human influence in some regions of the KAZA-TFCA originating from
farms, high human density, and road traffic.

@@

\subsection{GPS Relocation Data}
% Collection
We collected GPS relocation data of 16 wild dog coalitions (i.e. 7 female and 9
male coalitions) that dispersed from the core study area between 2011 and 2019.
Potential dispersing individuals were immobilized according to protocols
described in \cite{Osofsky.1996} and fitted with GPS/Satellite radio collars
(\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while they were still
with their natal pack. All required procedures were undertaken and supervised by
a Botswana-registered wildlife veterinarian. Immobilized individuals were
monitored and all were observed to successfully re-join their packs within one
hour after immobilization. GPS collars were programmed to record GPS relocation
data every 4 hours during dispersal and to regularly transmit them via Iridium
satellite system to a base station. The system thus allowed remote-tracking of
collared individuals, even where field conditions were prohibitive or dispersal
trajectories crossed international borders.

% Separation of Dispersers/Residents
Because we were mainly interested in recording dispersal trajectories, we
discarded any GPS data collected while individuals were still with their natal
packs \citep{Cozzi.2020}. Dispersal was deemed to have started when a coalition
had left its natal home range, and continued until it exhibited clear signs of
resident behavior. Some coalitions undertook exploratory movements prior to
proper dispersal. Such movements have been observed in other species and are
reported to be very similar to proper dispersal \citep{Killeen.2014}. We
therefore classified exploratory movements as dispersal too.

% Some Summary Statistics
In total, we collected 4'691 GPS relocations during dispersal (see
\Cref{Trajectories}). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}. On average,
dispersal coalitions included three individuals (min = 1, max = 7), dispersed
for 41 days (min = 7, max = 133), and covered a mean cumulative distance of 370
km (min = 105, max = 610).

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Trajectories.pdf}
    \caption{Illustration of the dispersal trajectories that we recorded. Each
    color represents a different dispersal coalition. All coalitions departed
    from our core study area, which is indicated by the white square. The
    coalition dispersing towards the far east of the map covered over 360 km in
    under 10 days.}
    \label{Trajectories}
  \end{center}
\end{figure}

\subsection{Spatial Covariates}
% General stuff
We used a set of geo-referenced covariates to investigate habitat preferences of
dispersing wild dogs (see \Cref{Covariates}). Covariates fell into one of the
three categories \textit{land cover}, \textit{protection status}, and
\textit{human influence}. Note that we did not include elevation as covariate
due to the absence of noteworthy elevational gradients in our study area. For
each covariate we obtained spatial raster layers from freely available online
services or from remote sensed satellite imagery. To ensure a consistent
resolution (i.e. cell-size or grain) across covariates, all layers were
coarsened or interpolated to match a resolution of 250m x 250m. Processing and
manipulation of data, as well as all spatial and statistical analyses were
performed in R, version 3.6.1 \citep{R.2019}.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Covariates.pdf}
    \caption{Overview of covariates that we used in our models. Although all
    covariates were prepared for the extended study area, we present only an
    extract to maximize visibility. The white square in each plot depicts our
    core study area. (a) Averaged layer of all binary maps of water and dryland
    that we acquired using our floodmapping algorithm. (b) Percentage cover of
    trees. (c) Percentage cover of shrubs/grassland. Note that anything not
    covered by trees or shrubs/grassland was deemed to be bareland. (d)
    Protection status according to the Peace Parks Foundation. (e) Human
    influence proxy, including human density, farms, and roads. (f) Distance to
    nearest road (white lines depict actual roads).}
    \label{Covariates}
  \end{center}
\end{figure}

\newpage
\subsubsection{Land Cover}
% Overview
Land cover was represented by four layers: a binary layer distinguishing water
and dryland, and three continuous layers showing the percentage cover by
different vegetation types, together adding up to 100\%.

% Water Layers
Water included open waters, rivers, wetlands, and swamps, whereas anything else
was presumed to be dryland. Because the flood extent of the Okavango Delta is
highly variable within and between years, we created weekly (8 days)
``floodmaps'', following the algorithm developed in \cite{Wolski.2017}. The
algorithm uses thresholding of SWIR reflectances of MODIS Terra satellite
imagery (MCD43A4, \cite{Schaaf.2015}) to separate water and dryland. As we
rewrote and slightly amended the original Python code in R, we describe the
underlying principles and indicate the discrepancies to the original algorithm
in \Cref{Appendix:FloodmappingAlgorithm}. While water in the Okavango Delta was
mapped dynamically, we used a static watermap for its surroundings. This static
map was based on Globeland's land cover dataset \citep{Chen.2015}, from which we
only retained the categories \textit{wetland} and \textit{waterbodies} and
collectively reclassified them to \textit{water}. Globeland had an original
resolution of 30m x 30m, so we coarsened the layer to 250m x 250m using the mode
of each 250m x 250m cell. Lastly, we improved river representation using the
MERIT Hydro dataset \citep{Yamazaki.2019}, from which we added all rivers with a
width of over 10m to our watermaps.

% Vegetation Layers
The three vegetation layers that we used in addition to the water/dryland maps
were derived from the MODIS Terra Vegetation Continuous Fields dataset (MOD44B,
\cite{Dimiceli.2015}). Combined, the three layers added up to 100\% and depicted
the percentage cover of tree-vegetation (henceforth trees), non-tree-vegetation
(henceforth shrubs/grassland), and non-vegetated (henceforth bareland). To
ensure that areas covered by the flood would get 0\% vegetation cover (i.e.
100\% bareland cover), we used our floodmap that aligned with the creation date
of the layers (30 March 2017) and manually defined pixels covered by the flood
as being 100\% bareland. The MODIS vegetation layers had a resolution of 250m x
250m, so we did not need to coarsen or interpolate them.

\subsubsection{Human Influence}
We defined human influence using a proxy that incorporated anthropogenic
pressures arising from (1) human density, (2) farming activities, and (3) roads.
Previous research revealed that veterinary fences have no influence on wild dog
movements \citep{Cozzi.2013b}, so we did not include them in our analysis.

% Human Density
(1) We obtained spatial human density estimates through Facebook's 30m x 30m
high resolution population density dataset \citep{Facebook.2019}. The dataset
contained some human density that arose from lodges and other touristic
activities that we did not want to depict as human pressures, so we manually
removed such regions from the data. The resulting layer was coarsened to 250m x
250m by summing up human densities within 250m x 250m cells.

% Farming
(2) Information on farming activities was sourced from the Globeland
\citep{Chen.2015} and Cropland \citep{Xiong.2017} land cover datasets, from
which we only retained pixels that were classified as either \textit{cultivated
land} or \textit{croplands}. Both layers had a resolution of 30m x 30m and were
coarsened to 250m x 250m by summing up values within 250m x 250m cells. The
layers were then merged into a single binary layer depicting presence (= 1) or
absence (= 0) of farms within 250m x 250m cells.

% Roads
(3) Geo-referenced data on roads was downloaded in shapefile format from Open
Street Map \citep{OpenStreetMap.2019}. We omitted minor roads and only retained
main tar roads and trunks belonging to the biggest four categories (see
\Cref{RoadsDescription} in \Cref{Appendix:RoadCategories}). Finally, we
rasterized roads and trunks onto a binary raster (1 = roads, 0 = no roads) with
250m x 250m resolution.
